---
title: "Distance Weighting Functions for Spatial Covariance Models"
author: "Guillaume Guénard"
date: "31/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Distance weighting functions

The the present study is concerned with the development of a spatial modelling
framework. In a nutshell, its consists in:

1. finding a function of the distance that is suitable to represent
the spatial covariance,

2. decompose these spatial covariance patterns as a set of eigenfunctions, and

3. use these eigenfunctions with ancillary representation
methods (e.g., linear models, generalized linear models, discriminant functions,
articifical neural networks).

Its purpose is to incorporate the information about where observations were
taken from (i.e., the spatial information) into analytical and predictive
models.

Distance weighting functions (DWF) describe relationships between the distances,
which are represented as matrix $\mathbf{D} = [d_{i,j}]$ of pairwise geographic
distance between the sites (or between the obserations and a set of
representative kernels), and their corresponding degree of association, which is
represented as matrix $\mathbf{W}=[w_{i,j}]$. Finding a suitable DWF is at the
basis of representing spatial patterns. The DWF is used to estimate the spatial
covariance (matrix $\mathbf{\Sigma}$, step #1 in the list above). In the
framework developed here, $\mathbf{W}$, obtained by applying the a DWF on
$\mathbf{D}$ in an element-wise manner. The complete process is described
in-length in the main text and involves centering $\mathbf{W}$ on its columns
(i.e., the resulting transformed columns have means of $0$) and multiplying it
by its own transpose. That last step garranties that the spatial covariance
matrix remains positive semi-definite, which is expected by the maximum
likelihood estimator (MLE). That estimator is, in turn, necessary to determine
which DWF to choose from and the most adequate values for its parameters. In in
developing the DWF, we have followed a few guidelines to help in obtaining
well-behaved MLE. A DWF:

1. has unit range on its entire domain, with the maximum weight being $1$ and
occurring when the distance is $0$, since $w(d=0)=1$, and the minimum weight
being $0$ and occurring when $d\geq \beta$ (i.e., $w(d\geq\beta)=0)$, where
$\beta$ is the parameter describing the maximum influence distance;

2. is continuous between $d=0$ and $d=\beta$;

3. is monotonically decreasing between $d=0$ and $d=\beta$;

4. converges into the same linear function $w=1-d/\beta$ when its shape
parameter $\alpha=1$ and into a Kronecker's $\delta$ (i.e., a function giving
the value $1$ for $d=0$ and $0$ everywhere else) when $\alpha=0$;

5. is initially steep, with its first derivatives with respect to $d$ having
values $\leq-1/\beta$ at $d=0$, which are then increasing to values
$\leq 0$ at $d=\beta$;

6. has a first derivatives with respect to $d$ that always increasing, i.e., its
second derivative with respect to $d$ has a value $\geq 0$ between $d=0$ and
$d=\beta$; and

7. has a parameter $\alpha$, driving its shape in the $0<\alpha<1$ interval.

Below, we describe five DWF. Some of them where loosely inspired by variogram
models used in kriging, whereas others are from standard analytic geometry, with
all the them complying with the properties enumerated above.

## Spherical function

We began by considering the spherical variogram model. It is a third order
polynomial which we adapted to our framework as follows:
$$w_{\beta}(d) = 1 - 1.5 \left(\frac{d}{\beta}\right) + 0.5\left(\frac{d}{\beta}\right)^3,$$
where $\beta$ corresponds to the range parameter. Its first derivative (with
respect to $d$) is $w'_{\beta}(d) = -1.5 + 1.5(d/\beta)^2$: the function's rate
of decrease is $-1.5$ at $d=0$ and increases to $0$ at $d=\beta$ in a quadratic
manner. To enable on to modulate that function's shape, let $a$ be the absolute
value of the initial rate of the function, which will now read as follows:
$$w_{a,\beta}(d) = 1 - a \left(\frac{d}{\beta}\right) + (a-1)\left(\frac{d}{\beta}\right)^{\frac{a}{a-1}}.$$
That function, which equal the one above for $a=1.5$ would comply with above
properties 1-3 for $a \in [-\infty,0]\cup[1,+\infty]$. Curves obtained by
varying $a$ from $1$ to $+\infty$ are continuous with those obtained when $a$
goes from $-\infty$ to $0$, as limits at $\pm\infty$ are identical and defined
as follow:
$$\lim_{a\rightarrow+\infty} w_{a,\beta}(d) = \lim_{a\rightarrow-\infty} w_{a,\beta}(d) = \left(\frac{d}{\beta}\right)\log\left(\frac{d}{\beta}\right) - \left(\frac{d}{\beta}\right) + 1.$$
We remapped the range of $a$ into the interval $[0,1]$ by representing it as an
rectangular hyperbolic function of $\alpha$ as follows:
$$a=\frac{1}{2}-\frac{1}{4\alpha-2}=\frac{\alpha-1}{2\alpha-1}.$$
The resulting function's first derivative is, by definition $-a/\beta$ at $d=0$
and $0$ at $d=\beta$ except when $a=0$, in which case it is $-1/\beta$ for all
$0\leq d\leq\beta$. The derivative of the function is:
$$w'_{a,\beta}(d) = \begin{cases} \dfrac{a\left(\frac{d}{\beta}\right)^\frac{a}{a-1}}{d}-\dfrac{a}{\beta}, & \alpha\in [0,1]\not\,\,\, \lbrace 0.5 \rbrace \\ \dfrac{\log\left(\frac{d}{\beta}\right)}{\beta}, & \alpha=0.5\,\cap\, d>0 \\ 1, & \alpha=0.5\,\cap\, d=0 \end{cases}.$$
In the particular case when $\alpha=0.25$ (i.e., $a=1.5$), the DWF corresponds to
the spherical variogram model. However, the former has more flexibility and can
represent a broader array of polynomial relationships.

```{r,echo=FALSE}
dst <- seq(0,5,0.01)
f <- function(d,alpha,beta) {
  dd <- d/beta
  if(alpha==0.5) {
    out <- dd * log(dd) - dd + 1
  } else {
    a <- 0.5 - 0.25/(0.5 - alpha)
    n <- a / (a - 1)
    out <- 1 - a*dd + (a - 1)*dd^n
  }
  out[d>beta] <- 0
  out
}
par(mar=c(5,5,2,2))
plot(y=f(dst,1,4),x=dst,type="l",ylim=c(0,1),xlab=expression(italic(d)),
     ylab=expression(italic(w)[list(spherical,alpha,beta)](italic(d))))
lines(y=f(dst,0.75,4),x=dst,lty=2L)
lines(y=f(dst,0.5,4),x=dst,lty=3L)
lines(y=f(dst,0.25,4),x=dst,lty=4L)
lines(y=f(dst,0.1,4),x=dst,lty=5L)
legend(x=3.5,y=1,lty=1L:5L,
       legend=parse(text=sprintf("alpha == %0.2f",c(1,0.75,0.5,0.25,0.1))))
```

Example of the spherical DWF for different values of $\alpha$ and $\beta=4$.

---

## Exponential function

The second function we considered was the exponential variogram function, which
we adapted to the framework described previously as follows:
$$w_{\beta}(d) = \dfrac{\mathrm{e}^{\frac{-3d}{\beta}}-\mathrm{e}^{-3}}{1-\mathrm{e}^{-3}}.$$
To enable one to modulate its shape, we began by replacing the value $-3$ by a
parameter $a$ as follows:
$$w_{a,\beta}(d) = \dfrac{\mathrm{e}^{\frac{ad}{\beta}}-\mathrm{e}^a}{1-\mathrm{e}^a},$$
with values of $a\in[-\infty,0]$ giving curves complying with above properties
1-3. By posing $a=(\alpha-1)/\alpha$, we obtained the following:
$$w_{\alpha,\beta}(d) = \dfrac{\mathrm{e}^{\frac{(\alpha-1)d}{\alpha\beta}}-\mathrm{e}^\frac{\alpha-1}{\alpha}}{1-\mathrm{e}^\frac{\alpha-1}{\alpha}},$$
which comply with properties 5-6. The first derivative of that function with
respect to $d$ is:
$$w'_{\alpha,\beta}(d) = \dfrac{\left(\alpha-1\right)\mathrm{e}^\frac{\left(\alpha-1\right)d}{\alpha \beta}}{\alpha\beta\left(1-\mathrm{e}^\frac{\alpha-1}{\alpha}\right)}.$$
That function can be used to help in interpreting DWF parameters fit to
empirical data.

```{r,echo=FALSE}
dst <- seq(0,5,0.01)
f <- function(d,alpha,beta) {
  if(alpha<1) {
        if(alpha>0) {
          a <- (alpha-1)/alpha
          out <- (exp(a*(d/beta)) - exp(a))/(1 - exp(a))
        } else {
          out <- numeric(length(d))
          out[d==0] <- 1
        }
      } else {
        out <- 1-d/beta
      }
      out[d>beta] <- 0
      out
}
par(mar=c(5,5,2,2))
plot(y=f(dst,1,4),x=dst,type="l",ylim=c(0,1),xlab=expression(italic(d)),
     ylab=expression(italic(w)[list(exponential,alpha,beta)](italic(d))))
lines(y=f(dst,0.75,4),x=dst,lty=2L)
lines(y=f(dst,0.5,4),x=dst,lty=3L)
lines(y=f(dst,0.25,4),x=dst,lty=4L)
lines(y=f(dst,0.1,4),x=dst,lty=5L)
legend(x=3.5,y=1,lty=1L:5L,
       legend=parse(text=sprintf("alpha == %0.2f",c(1,0.75,0.5,0.25,0.1))))
```

Example of the exponential power weighting for different values of $\alpha$ and
$\beta=4$.

---

We did not seek to develop DWF on the basis of the Gaussian and hole effect
variograms models. The Gaussian variogram model has a slope that is consistently
above $-1/\beta$ and is not adaptable to the present framework. The hole effect
variogram model, which is based on the canonical sine function, is not
consistently decreasing. The next weighting function were developed from general
analytic geometric functions that are not commonly used as variogram models.

## Power function

We began the definitions of the third weighting function, which we called the
«power» weighting function, by considering a simple linear function of the form
$y = ax+b$, with $a = -\frac{1}{d_{max}}$ and $b = 1$ to comply with the
properties mentioned previously, and raised it to the inverse of $\alpha$ as
follows:

$$w_{\alpha,\beta}(d) = \left(1-\dfrac{d}{\beta}\right)^\frac{1}{\alpha},$$

with its first derivative with respect to $d$ being:

$$w'_{\alpha,\beta}(d) = -\dfrac{\left(1-\frac{d}{\beta}\right)^{\frac{1}{\alpha}-1}}{\beta\alpha}.$$

Value of the derivative at limit $d=0$ is thus $-\frac{1}{\beta\alpha}$ for
all $\alpha \in ]0,1]$ whereas at limit $d=\beta$ it is $-\frac{1}{\beta}$
whenever $\alpha=1$ and $0$ for all $\alpha \in ]0,1[$.

As for all above-defined DWF, the power weighting function takes the shape of a
straight line when $\alpha = 1$ whereas for other non-zero $a<1$ it takes the
shape of curves starting steeply (the smaller the $\alpha$, the steeper the
curve starts) and ending gently (the smaller the $\alpha$ the gentler the curve
ends).

```{r,echo=FALSE}
dst <- seq(0,5,0.01)
f <- function(d,alpha,beta) {
  out <- (1-d/beta)^(1/alpha)
  out[d>beta] <- 0
  out
}
par(mar=c(5,5,2,2))
plot(y=f(dst,1,4),x=dst,type="l",ylim=c(0,1),xlab=expression(italic(d)),
     ylab=expression(italic(w)[list(power,alpha,beta)](italic(d))))
lines(y=f(dst,0.75,4),x=dst,lty=2L)
lines(y=f(dst,0.5,4),x=dst,lty=3L)
lines(y=f(dst,0.25,4),x=dst,lty=4L)
lines(y=f(dst,0.1,4),x=dst,lty=5L)
legend(x=3.5,y=1,lty=1L:5L,
       legend=parse(text=sprintf("alpha == %0.2f",c(1,0.75,0.5,0.25,0.1))))
```

Example of the power weighting for different values of $\alpha$ and $\beta=4$.

---

## Hyperbolic weighting function

The fourth DWF was obtained using a rectangular hyperbola as its backbone and
was named accordingly the «hyperbolic» weighting function. The rectangular
hyperbola, $y = 1/x$, was horizontally offset it by one to the left to obtain a
function taking the value $w=1$ at $d=0$ and raise it to the power $1/a$ to
apply control over its shape. The resulting function had the following form
$y=(x+1)^{-\frac{1}{a}}$, which, at $d=\beta$ take the value
$(\beta+1)^{-\frac{1}{a}}$. We thus vertically offset the function downward by
that value and scale the resulting function within the $[0,1]$ interval,
thereby obtaining the following:
$$w_{a,\beta}(d)= \dfrac{\left(d+1\right)^{-\frac{1}{a}}-\left(\beta+1\right)^{-\frac{1}{a}}}{1-\left(\beta+1\right)^{-\frac{1}{a}}}.$$
Similarly as for the spherical weighting function presented earlier, the
function is well-behaved for a disjoint interval, this time for
$-a \in [-\infty,0]\cup[1,+\infty]$. We solve that issue in a similar manner,
as for the spherical DWF by remapping the range of $-a$ into the interval
$[0,1]$ by representing it as an other rectangular hyperbolic function of
$\alpha$ as follows:
$$a=\frac{1}{2-4\alpha}-\frac{1}{2}=\frac{\alpha}{1-2\alpha},$$
with the resulting weighting function reading as follows:
$$w_{\alpha,\beta}(d)= \dfrac{\left(d+1\right)^{2-\frac{1}{\alpha}}-\left(\beta+1\right)^{2-\frac{1}{\alpha}}}{1-\left(\beta+1\right)^{2-\frac{1}{\alpha}}}.$$
For $\alpha=0.5$ (and thus $a=\pm\infty$) the function becomes:
$$w_{\alpha=0.5,\beta}(d) = \lim_{a\rightarrow+\infty} w_{a,\beta}(d) = \lim_{a\rightarrow-\infty} w_{a,\beta}(d) = 1-\dfrac{\log(d+1)}{\log(\beta+1)}$$



and with its first derivative with respect to $d$ being:

$$w'_{d_\alpha,\beta}(d)=\dfrac{\left(2-\frac{1}{\alpha}\right)\left(d+1\right)^{1-\frac{1}{\alpha}}}{1-\left(\beta+1\right)^{2-\frac{1}{\alpha}}}$$
for $\alpha\in [0,1]\not\,\,\lbrace 0.5 \rbrace$ and:
$$w'_{d_\alpha,\beta}(d)=-\dfrac{1}{\log(\beta+1)(d+1)}$$
for $\alpha=0.5$. The function features a steep initial slope and a gentle final
slope, with initial steepness (and final gentleness) being more pronounced for
small values of $\alpha$. As for the spherical and power weighting functions, it
amounts to the linear function $w=1-d/\beta$ when $\alpha=1$.

```{r,echo=FALSE}
dst <- seq(0,5,0.01)
f <- function(d,alpha,beta) {
  if(alpha == 0.5) {
    out <- 1 - log(d + 1)/log(beta + 1)
  } else {
    a <- 0.5 - 0.25/(0.5 - alpha)
    out <- ((1+d)^(1/a) - (1+beta)^(1/a))/(1 - (1+beta)^(1/a))
  }
  out[d>beta] <- 0
  out
}
par(mar=c(5,5,2,2))
plot(y=f(dst,1,4),x=dst,type="l",ylim=c(0,1),xlab=expression(italic(d)),
     ylab=expression(italic(w)[list(hyperbolic,alpha,beta)](italic(d))))
lines(y=f(dst,0.75,4),x=dst,lty=2L)
lines(y=f(dst,0.5,4),x=dst,lty=3L)
lines(y=f(dst,0.25,4),x=dst,lty=4L)
lines(y=f(dst,0.1,4),x=dst,lty=5L)
legend(x=3.5,y=1,lty=1L:5L,
       legend=parse(text=sprintf("alpha == %0.2f",c(1,0.75,0.5,0.25,0.1))))
```

Example of the hyperbolic weighting for different values of $\alpha$ and
$\beta=4$.

---

## Superelliptical weighting function

Finally, we called the fifth and last weighting function «superelliptical»
weighting function as it is based on superelliptical (also known as the Lamé)
curves, a generalization the equation of an ellipse. As such, its first
derivative with respect to $d$ has the value $-\infty$ at $d=0$ and $0$ at
$d=\beta$. It is defined as follows:
$$w_{\alpha,\beta}(d) = 1-\dfrac{\left(\beta^\frac{1}{\alpha}-\left(\beta-d\right)^\frac{1}{\alpha}\right)^\alpha}{\beta}$$

and corresponds exactly to the lower-left portion of a vertically-compressed
ellipse going through point $(d=0,w=1)$ and $(d=\beta,w=0)$ when $\alpha=0.5$.
Its first derivative with respect to $d$ is:

$$w'_{\alpha,\beta}(d) = -\dfrac{\left(\beta^\frac{1}{\alpha}-\left(\beta-d\right)^\frac{1}{\alpha}\right)^{\alpha-1}\left(\beta-d\right)^{\frac{1}{\alpha}-1}}{\beta}.$$

The shape of the function becomes increasingly rectangular as one decreases
$\alpha$ from $0.5$ toward $0$ and increasingly linear as one increases $\alpha$
from $0.5$ toward $1$. As for the spherical, hyperbolic, and power weighting
function, the superelliptical weighting function becomes a straight line when
$\alpha=1$.

```{r,echo=FALSE}
dst <- seq(0,5,0.01)
f <- function(d,alpha,beta) {
  out <- 1 - (1/beta)*(beta^(1/alpha) - (beta-d)^(1/alpha))^alpha
  out[d>beta] <- 0
  out
}
par(mar=c(5,5,2,2))
plot(y=f(dst,1,4),x=dst,type="l",ylim=c(0,1),xlab=expression(italic(d)),
     ylab=expression(italic(w)[list(superelliptical,alpha,beta)](italic(d))))
lines(y=f(dst,0.75,4),x=dst,lty=2L)
lines(y=f(dst,0.5,4),x=dst,lty=3L)
lines(y=f(dst,0.25,4),x=dst,lty=4L)
lines(y=f(dst,0.1,4),x=dst,lty=5L)
legend(x=3.5,y=1,lty=1L:5L,
       legend=parse(text=sprintf("alpha == %0.2f",c(1,0.75,0.5,0.25,0.1))))
```

Example of the superelliptical weighting for different values of $\alpha$ and
$\beta=4$.

---

Although it is our hope that the five DWF presented here cover a broad array of
possible distance-weights relationships, it is by no mean an exhaustive listing
of all the possible relationships that can be used for modelling spatial
covariance from among-site distances. It is also our hope that scientists supply
any lack of a relevant function that may arise in the future.
